Converging High-Throughput and High-Performance Computing: A Case Study

### 2017-04-04 PandaEvolution.tex
saved 4-5 lines and minor corrections.
Scattered space saving editings

The next LHC run for taking data (Run 3) will require more resources than the
Worldwide LHC Computing Grid (WLCG) can provide. Currently, PanDA WMS uses more
than 300,000 cores at over 100 Grid sites, with a peak performance of 0.3
petaFLOPS. This capacity will be sufficient for the planned analysis and data
processing, but it will be insufficient for the Monte Carlo production workflow
and any extra activity. To alleviate these challenges, ATLAS is engaged in a
program to expand the current computing model to include additional resources
such as the opportunistic use of supercomputers.

Modern supercomputers have been designed mainly to support parallel computation
that requires runtime communication. Job execution is parallelized across
multiple cores, each core calculating a small part of the problem and
communicating with other cores via a message passing interface (MPI).
Accordingly, supercomputers have large number of worker nodes, connected through
a high-speed, low-latency dedicated network. Each worker node has multicore
CPUs, usually augmented with parallel Graphics Processing Units (GPUs) or other
types of specialized coprocessors.

@@

Titan's users submit jobs to Titan's PBS scheduler by logging into login or data
transfer nodes (DTNs). Titan's authentication and authorization model is based
on two-factor authentication with a RSA SecurID key, generated every 30 seconds.
Login nodes and DTNs have out/inbound wide area network connectivity while
worker nodes have only local network access. Fair-share and allocation policies
are in place both for the PBS batch system and shared file systems.

@@

Currently, very few HEP applications can benefit from Titan's GPUs but some
computationally-intensive and non memory-intensive tasks of ATLAS workflows can
be off-loaded from the Grid to Titan's large amount of cores. Further, when HEP
tasks can be partitioned into independent jobs, Titan worker nodes can be used
to execute up to 16 concurrent payloads, one per each available core. Given
these constraints and challenges, the type of task most suitable for execution
at the moment on Titan is Monte Carlo detector simulation. This type of task is
mostly computational-intensive, requiring less than 2GB of RAM at runtime and
with small input data requirements. Detector simulation tasks in ATLAS are
performed via AthenaMP~\cite{aad2010atlas}, the ATLAS software framework
integrating the GEANT4 simulation toolkit~\cite{agostinelli2003geant4}. These
tasks account for $\approx$ 60\% of all the jobs on WLCG, making them a primary
candidate for offloading.

Detector simulation is part of the ATLAS production Monte Carlo (MC) workflow
(also known as MC production
chain)~\cite{rimoldi2006atlas,de2013delphes,ritsch2014atlas}. The MC workflow
consists of four main stages: event generation, detector simulation,
digitization, and reconstruction. Event generation creates sets of particle
four-momenta via different generators, e.g., PYTHIA~\cite{sjostrand2006pythia},
HERWIG~\cite{corcella2001herwig} and many others. The detector simulator is
called Geant4~\cite{agostinelli2003geant4} and simulates the ATLAS detector and
the interaction between that detector and particles. Each interaction creates a
so-called hit and all hits are collected and passed on for digitalization, where
hits are further processed to mimic the readout of the detector. Finally,
reconstruction operates local pattern recognition, creating high-level objects
like particles and jets.

@@

The backfill optimization of Titan's Moab scheduler offers the opportunity to
avoid the overhead of queue wait times without using pilot
abstraction~\cite{maui_backfill_url}. This optimization enables Moab to start
low-priority jobs as long as they do not delay higher priority jobs,
independent of whether the low-priority jobs were queued after the
high-priority jobs.

@@

Compared to pilots, backfill has the disadvantage of limiting the amount of
worker nodes that can be requested. Pilots are normal jobs: they can request as
many worker nodes for as much time as a queue can offer. On the contrary, jobs
sized on the basis of an available backfill slot depend on the number of worker
nodes that cannot be given to any other job in the Titan's queue at that moment
in time.

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figures/panda_broker_architecture.pdf}
    \caption{PanDA Broker architecture as deployed on Titan. Numbers indicates
    the execution process of a detector simulation job, part of the production
    ATLAS MD workflow.}
\label{fig:panda_broker}
\end{figure}

@@

>>>>

The upcoming LHC Run 3 will require more resources than the Worldwide LHC
Computing Grid (WLCG) can provide. Currently, PanDA WMS uses more than 300,000
cores at over 100 Grid sites, with a peak performance of 0.3 petaFLOPS. This
capacity will be sufficient for the planned analysis and data processing, but it
will be insufficient for the Monte Carlo production workflow and any extra
activity. To alleviate these challenges, ATLAS is engaged in a program to expand
the current computing model to include additional resources such as the
opportunistic use of supercomputers.

Generally, supercomputers are designed to support parallel computation that
requires runtime communication. Jobs are executed across multiple cores, each
core calculating a small part of the problem and communicating with other cores
via MPI. Accordingly, supercomputers have large number of worker nodes,
connected through a high-speed, low-latency dedicated network. Each worker node
has multicore CPUs, usually augmented with Graphics Processing Units (GPUs) or
other specialized coprocessors.

@@

Titan's users submit jobs to Titan's PBS scheduler by logging into login or data
transfer nodes (DTNs). Titan's authentication and authorization model is based
on two-factor authentication with a RSA SecurID key. Login nodes and DTNs have
out/inbound wide area network connectivity while worker nodes have only local
network access. Fair-share and allocation policies are in place both for the PBS
batch system and shared file systems.

@@

Currently, very few HEP applications can benefit from Titan's GPUs but some
computationally-intensive and non memory-intensive tasks of ATLAS workflows can
be off-loaded from the Grid to Titan's. Further, when HEP tasks can be
partitioned into independent jobs, Titan worker nodes can be used to execute up
to 16 concurrent payloads, one per each available core. Given these constraints
and challenges, the type of task most suitable for execution at the moment on
Titan is Monte Carlo detector simulation. This type of task is mostly
computational-intensive, requiring less than 2GB of RAM at runtime and with
small input data requirements. Detector simulation tasks in ATLAS are performed
via AthenaMP~\cite{aad2010atlas}, the ATLAS software framework integrating the
GEANT4 simulation toolkit~\cite{agostinelli2003geant4}. These tasks account for
$\approx$ 60\% of all the jobs on WLCG, making them a primary candidate for
offloading.

Detector simulation is part of the ATLAS production Monte Carlo (MC)
workflow~\cite{rimoldi2006atlas,de2013delphes,ritsch2014atlas}. The MC workflow
consists of four main stages: event generation, detector simulation,
digitization, and reconstruction. Event generation creates sets of particle
four-momenta via different generators, e.g., PYTHIA~\cite{sjostrand2006pythia},
HERWIG~\cite{corcella2001herwig} and many others. The detector simulator is
called Geant4~\cite{agostinelli2003geant4} and simulates the ATLAS detector and
the interaction between that detector and particles. Each interaction creates a
so-called hit and all hits are collected and passed on for digitalization, where
hits are further processed to mimic the readout of the detector. Finally,
reconstruction operates local pattern recognition, creating high-level objects
like particles and jets.

@@

The backfill optimization of Titan's Moab scheduler allows to avoid the overhead
of queue wait times without using pilot abstraction~\cite{maui_backfill_url}.
With this optimization, Moab starts low-priority jobs when they do not delay
higher priority jobs, independent of whether the low-priority jobs were queued
after the high-priority jobs.

@@

Compared to pilots, backfill has the disadvantage of limiting the amount of
worker nodes that can be requested. Pilots are normal jobs: they can request as
many worker nodes and walltime as a queue can offer. On the contrary, jobs sized
according to an available backfill slot depend on the number of worker nodes and
walltime that cannot be given to any other job at that moment in time.

@@

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figures/panda_broker_architecture.pdf}
    \caption{PanDA Broker architecture as deployed on Titan. Numbers indicates
    the execution process of a detector simulation job, part of the production
    ATLAS MC workflow.}
\label{fig:panda_broker}
\end{figure}
